Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-13 paper-ieee-trans.tex

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
supporting an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, motivated
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the 5th, 6th and 7th Framework Programmes,
and recently by the Horizon 2020 programme \cite{h2020} -- are under way to
develop software solutions that provide a seamless and transparent access to distributed
computing and data resources by the researchers.

@@

The {\sl gLite} middleware \cite{glite} was the
official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
re--engineering efforts to converge with the {\sl LHC} Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, which issued more than 10 thousand bug, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of operational tools, such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
management of the software provisioning process \cite{mario}, to validate and distribute the software to
the production infrastructure. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defines a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above with the additional complexity of being framed into
large collaborative projects with an extensive list of partners.

In the following sections, we will describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition}, will provide guidelines to develop quality
software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequacy in terms of scope, objective fulfilment,
documentation completeness, etc. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

The practical implementation of the CI scenario leveraged on tightly integrated open
source tools such as GitHub \cite{github} -- as the online source code repository --,
Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually more correlated. The software updates of
{\sl INDIGO--DataCloud} products, taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI-validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the {\sl INDIGO--DataCloud}’s application repository.

@@

After more that a decade of experiences collected throughout the reviewed
EU-funded software development and e-Infrastructure management projects, it
becomes apparent that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. The gradual adoption of software engineering insights, such as
DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
collaborative tools, extending their capabilities to bring along tight
integrations amongst other services, matched the requirements of such practices
and procedures, facilitating in a great manner their implementation in real scenarios.

Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre-production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process --, alleviates the likelihood of discovering uncovered
issues at production.

>>>>

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
supporting an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Grid computing infrastructure, motivated
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the 5th, 6th and 7th Framework Programmes,
and recently by the Horizon 2020 programme \cite{h2020} -- are under way to
develop software solutions that provide a seamless and transparent access to distributed
computing and data resources by the researchers.

@@

The {\sl gLite} middleware \cite{glite} was the
official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
re--engineering efforts to converge with the {\sl LHC} Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
development team was comprised of more than 80 people from 12 academic and
industrial partners, which issued more than 10 thousand bug, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of operational tools, such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid Operations Configuration DataBase (GOCDB) \cite{gocdb}, and the
management of the software provisioning process \cite{mario}, to validate and distribute the software to
the production infrastructure. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers, such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. QC defines a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above achievements were with the additional complexity of being framed into
large collaborative projects with an extensive list of partners.

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} provide guidelines to develop quality
software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} promotes automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
and documentation completeness. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

The practical implementation of the CI scenario leveraged on tightly integrated open
source tools such as GitHub \cite{github} as the online source code repository,
Jenkins \cite{jenkins}, the event-response CI system, and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually are more correlated. The software updates of
{\sl INDIGO--DataCloud} products, taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI-validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the {\sl INDIGO--DataCloud}'s application repository.

@@

After more than a decade of experiences collected throughout the reviewed
EU-funded software development and e-Infrastructure management projects, it
becomes apparent that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. The gradual adoption of software engineering insights, such as
DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
collaborative tools, extending their capabilities to bring along tight
integrations amongst other services, matched the requirements of such practices
and procedures, facilitating in a great manner their implementation in real scenarios.

Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre-production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.
